---
title: "performr vignette"
author: "Silas Tittes"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

How to use `performr` package
===============

`performr` implements a probabilistic bayesian hierarchical model (implemented in (Stan)[http://mc-stan.org/]) to predict performance curves for a set of input taxa. The manuscript that describes the method is in revision. This page is intended to help those interested in using the model.

Installation
---------------

Installing `performr` should be done using the following command:
```{r, eval=FALSE}
#install
devtools::install_github("silastittes/performr")

#load library
library(performr)
```

The `performr` depends on the `tidyverse`, which can take some time to install. Make sure your R is up to date! 

Simulating example data 
---------------

```{r}
simulate_data()
```


Fitting the model with `performr::stan_performance()`
---------------

Output
---------------


Posterior predictive data generation
---------------

